---
title: "Spatial Operations"
format: html
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,collapse=TRUE)
```

```{r}

library(sf)      # vector data package
library(terra)   # raster data package
library(dplyr)   # tidyverse package for data frame manipulation
library(spData)  # loads datasets used here
library(here)
library(tmap)
```

-   We will use the following datasets.

```{r}
elev = rast(system.file("raster/elev.tif", package = "spData"))
grain = rast(system.file("raster/grain.tif", package = "spData"))
```

# Introduction

<div>

There are three tasks that we will cover today

1.  Spatial filtering

    <div>

    -   With spatial queries like: Does it intersects?

    </div>

2.  Spatial Joins

    <div>

    -   Spatial joins between vector datasets

    </div>

3.  Aggregation.

    <div>

    -   Grouping by geometry

    </div>

</div>

<div>

**There is a link between attribute operations and spatial ones:**

-   spatial subsetting: select rows based on geom.

-   spatial joining: combine tables based on geom.

-   aggregation: group observation based on geom.

</div>

<div>

-   Spatial joins, for example, can be done in a number of ways:

    -   matching entities that intersect with or are close enough to the target spot.

-   To explore the spatial relationships (contained, overlaps, etc.) between objects:

    -   use functions (**topological relations**) on sf objects.

-   Distances: all spatial objects are related through space.

    -   Distance calculations can be used to explore the strength of this relationship.

</div>

[**With respect to Rasters**]{.underline}

<div>

-   Spatial operations on raster objects include subsetting and merging several raster 'tiles' into a single object.

-   Map algebra covers a range of operations that modify raster cell values, with or without reference to surrounding cell values

    -   vital for many applications.

-   We will also compute distances within rasters.

-   Note that to apply any function on two spatial objects, the latter most share the same CRS!

</div>

## Vector data: subsetting

<div>

-   **Goal:** reshape an existing object in reference to another object.

-   Subsets of `sf` data frames can be created with **square bracket (\[)** operator.

    -   Syntax `x[y, , op = st_intersects]`, where `x` is an `sf` object from which a subset of rows will be returned.
    -   `y` is the 'subsetting object' `op = st_intersects` specifies the topological relation to do the subsetting.

</div>

<div>

-   The **default** topological relation is st_intersects()

    -   the command x\[y, \] is identical to x\[y, , op = st_intersects\]

-   The `filter()` function from the `tidyverse` can also be used.

</div>

## Vector data: subsetting

-   Demonstration: `nz` and `nz_height` datasets.

    -   contain geographic data on the 16 main regions and 101 highest points in New Zealand (projected CRS).

-   Create an object representing Canterbury and return all high points in the region:

```{r}

# filter out Canterbury
canterbury = nz %>% filter(Name == "Canterbury")
# subset the high points that "intersect" the above, use nz_height.
canterbury_height = nz_height[canterbury,, op = st_intersects]
```

```{r}

tm_shape(canterbury) + 
  tm_borders() +
  tm_shape(canterbury_height) +
  tm_symbols(shape = 17, col = "blue", size = .2)


```

**In summary:**

<div>

-   The command `x[y, ]` subsets features of a **target** `x` w.r.t. object `y`.

-   Both `x` and `y` must be geographic objects (`sf`).

-   Various topological relations for subsetting:

    -   `touches`, `crosses` or `within` (among others).

-   `st_intersects` is a **'catch all' instruction**

    -   catches everything that touches, crosses or falls within the source 'subsetting' object

</div>

<div>

-   Alternative spatial operators:

    -   the opposite to st_intersects:
    -   `nz_height[canterbury, op = st_disjoint]`

-   plot the map of New Zealand and the high points [outside Canterbury]{.underline}.

    -   `nz_height[canterbury, op = st_disjoint]`

</div>

[SUBSETTING in smaller steps]{.underline}

<div>

**Topological Operators Outputs**

-   They return [**sparse**]{.underline} objects that can be used for subsetting.

-   In the below code, we create an object with `(empty)` and `1`.

    <div>

    -   `empty` indicates no intersection between the target object and the subsetting object.

    </div>

-   it is an empty vector with length zero.

-   Then we transform the latter into a **logical vector**.

-   Finally we conduct the **subsetting operation**.

</div>

------------------------------------------------------------------------

```{r}

# intersect heights and Canterbury
heights_in_cant_sparse = st_intersects(x = nz_height, y = canterbury)

class(heights_in_cant_sparse)

heights_in_cant_sparse

# transform this into a logical
# lengths: applied to each element in a list

heights_in_cant_logical = lengths(heights_in_cant_sparse) > 0

# carry out the subsetting

canterbury_height2 = nz_height[heights_in_cant_logical, ]
```

**THE dplyr approach:**

-   `st_filter`: similar to the standard `dplyr`.

```{r}
canterbury_height3 = nz_height %>%
  st_filter(y = canterbury, .predicate = st_intersects)
```

------------------------------------------------------------------------

# Vector data: spatial relations

-   Sometimes it is important to establish whether two objects are spatially related.

    -   **Topological relations**: pin down the existence of a spatial relation.

-   Symmetric operators:

1.  equals
2.  intersects
3.  crosses
4.  touches
5.  overlaps

-   Asymmetric operators:

1.  contains
2.  within

## **Visualization**

-   Let's create an example.
-   First, **we create a polygon**: use `cbind` to generate a matrix of vertices.
-   use `st_sfc` and `st_polygon` to create an `sf`.
-   we will **create a line and group of points**.
-   we will **visually** examine the spatial relationships.
-   Finally, we will use the **operators** (binary predicates) to corroborate our visual inspection.

```{r fig1, fig.asp= .5}

# A matrix has vectors inide:

polygon_matrix = cbind(
  x = c(0, 0, 1, 1,   0),
  y = c(0, 1, 1, 0.5, 0)
)

# then we need to set it as a list and classify it as a polygon:

polygon_matrix = list(polygon_matrix) %>% st_polygon()

# transform it into a sf-column

polygon_sfc = st_sfc(polygon_matrix)

tmap::tm_shape(polygon_sfc) + tmap::tm_polygons() + tmap::tm_grid(lines = FALSE)

```

Now let's do the line:

```{r}
line_matrix = cbind(
  x = c(0.4, 1),
  y = c(0.2, 0.5))

line_sfc = st_sfc(st_linestring(line_matrix))

```

Now a set of points:

-   To do this, you need data frame listing the x and y coordinates.

-   then, you need to apply st_as_sf indicating the coordinate columns.

```{r}

# create a data frame of points
(point_df = data.frame(
  coord_x = c(0.2, 0.7, 0.4),
  coord_y = c(0.1, 0.2, 0.8)
)) 
library(tibble)

point_sf = st_as_sf(point_df, coords = c("coord_x", "coord_y")) %>% 
  rowid_to_column("ID") %>% 
  mutate(ID=as.character(ID))

```

```{r fig2, fig.asp= .5}

oldw <- getOption("warn")
options(warn = -1)

tmap::tm_shape(polygon_sfc) + tmap::tm_polygons() + 
  tmap::tm_shape(line_sfc) + tmap::tm_lines(scale = 10) +
  tmap::tm_shape(point_sf) + 
  tmap::tm_dots(scale=5, legend.show = F,col = "turquoise", alpha = .7, size= .1) +
  tmap::tm_text("ID", size = .5) +
  tmap::tm_grid(lines = FALSE)

options(warn = oldw)

```

## Spatial Queries

-   Let's conduct a simple **query**.
-   Which of the points in `point_sf` intersect in some way with `polygon_sfc`?
-   This question can be answered with the spatial predicate `st_intersects()` as follows:

```{r}

# The code below sets sparse=FALSE to coerce the output 
# into a logical vector, instead of a sparse matrix.
#intercept boll ceviriyo alan içindeyse true
st_intersects(point_sf, polygon_sfc, sparse = FALSE)

# A sparse matrix is a list of vectors with
# empty elements where a match doe not exists.

```

-   Which points **lie within** the polygon?
-   Which features **are on or contain a shared boundary** with y?
-   These can be answered as follows:

```{r}
st_within(point_sf, polygon_sfc)
st_touches(point_sf, polygon_sfc)
```

-   The opposite of `st_intersects()` is `st_disjoint()`, which returns only objects that do not spatially relate in any way to the selecting object

```{r}

# note [, 1] converts the result into a vector:

st_disjoint(point_sf, polygon_sfc, sparse = FALSE)[, 1]

```

**Distance-based Queries**

-   `st_is_within_distance()` detects features within a distance from the target.

-   It can be used to set **how close target objects need to be** before they are selected.

-   Distances between points and polyongs are measured to the **nearest edge**,

    -   In our case **point 2's distance to the polygon** is measured with respect to the edge above it..
    -   Verify the actual distance between **point 2** and the polygon is 0.13 with the command `st_distance(point_sf, polygon_sfc)`.

-   The **'is within distance'** binary spatial predicate is demonstrated in the code chunk below,

-   Indeed, every point is within 0.2 units of the polygon:

```{r}
st_is_within_distance(point_sf, polygon_sfc,
                      dist = 0.1, sparse = FALSE)[, 1]

st_distance(point_sf,polygon_sfc)

```

------------------------------------------------------------------------

## Vector data: spatial joining

<div>

-   Joining two **non-spatial** datasets relies on a **shared 'key' variable**

-   Spatial data joining applies the same concept drawing **on spatial relations**

-   Joining adds new columns to the target object **x**, from a source object **y**.

-   *Example*:

    -   ten points randomly distributed across the Earth's surface
    -   for the points that are on land, which countries are they in?

[**MORE SPECIFICALLY**]{.underline}

-   Establish the `bbox` for the analysis: "the entire globe"

-   Create points that are randomly scattered over the Earth's surface. Use the r's uniform distribution, and make sure the values fall into the `bbox`

-   Set the points as an sf object.

</div>

```{r}
set.seed(2018) # set seed for reproducibility
(bb = st_bbox(world)) # the world's bounds

# create a data frame

random_df = data.frame(
  
  x = runif(n = 10, # number of points
            min = bb[1], # lower bound
            max = bb[3]),# upper bound
  
  y = runif(n = 10,
            min = bb[2],
            max = bb[4])
)

random_points = random_df %>% 
  st_as_sf(coords = c("x", "y")) %>% # set coordinates
  st_set_crs("EPSG:4326") # set geographic CRS

```

4.  Now, plot the points on an earth's map.

```{r fig5, fig.asp= .5}

st_crs(world) <- 4326

tm_shape(world) +
  tm_borders() +
  tm_shape(random_points) +
  tm_symbols(shape = 17, col = "blue", size = .2)


```

-   We want the object `world_random` yielding only countries that contain random points

    -   we will obtain it via a **spatial join**.

```{r}

# find the countries "touched" by random points
(world_random = world[random_points,])

```

-   Take the data_frame with the random points.
-   For each random point, find what country it belongs to.
-   `st_join` is the **crucial** function here.

```{r}
# find the points that touch a country.
(random_joined = 
   st_join(random_points, select(world,name_long),
           join = st_intersects))
```

```{r}
tm_shape(world) +
  tm_borders() +
  tm_shape(world_random) + 
  tm_polygons("name_long") + # This will set a different color to the selected countries
  tm_shape(random_points) + 
  tm_symbols(shape = 17, col = "blue", size = .2)

```

## a few comments on spatial joins.

<div>

-   By default, `st_join()` performs a left join

-   ALL ROWS from x including rows with no match in y.

-   It can also do [**inner joins**]{.underline}

    -   set the argument left = FALSE.

</div>

<div>

-   The default topological operator used by `st_join()` is `st_intersects()`

-   [Warning]{.underline}:

    -   When **x** contains polygons, each of which match multiple objects in **y**, spatial joins will result in duplicate features, creates a new row for each match in y (**see the homework**).

</div>

## Non-overlapping joins

-   Sometimes two geographic datasets do not touch but still have a strong geographic relationship.
-   The datasets `cycle_hire` and `cycle_hire_osm` provide a good example.
-   Plotting them shows that they are often closely related but they do not touch.

### London bike hire key information

-   You can hire bikes using London's public cycle hire scheme, Santander Cycles.
-   Riders will find 800 docking stations and 12,000 bikes to hire around London.
-   Bikes can be hired using a bank card at the docking station or using the official Santander Cycles app.

```{r}
plot(st_geometry(cycle_hire), col = "blue", main = "London Cycle points: official-blue, OpenStreetMap-red")
plot(st_geometry(cycle_hire_osm), add = TRUE, pch = 3, col = "red")
```

-   We can check if any points are the same:

    -   `any`: given a set of logical vectors, is at least one of the values true?

```{r}
st_touches(cycle_hire, cycle_hire_osm, sparse = FALSE) %>% 
any()

```

<div>

**THE TASK AT HAND**

-   Imagine that we need to join the **capacity variable** in `cycle_hire_osm` onto the official 'target' data contained in `cycle_hire`.
-   This is when a non-overlapping join is needed.
-   The simplest method is to use the topological operator st_is_within_distance()
    -   use a threshold distance of 20 m.
    -   [that is, assume that if two points, belonging each to a different dataset, are close enough, then they speak about the same spot]{.underline}.

</div>

```{r}
head(cycle_hire)
```

```{r}
head(cycle_hire_osm)
```

-   How to retrieve the values associated with the respective `cycle_hire_osm` points?
-   The solution is again with `st_join()`.

```{r}

aux = st_join(cycle_hire,
              select(cycle_hire_osm,capacity),
              join = st_is_within_distance,
            dist = 20)

nrow(cycle_hire)
nrow(aux)
head(aux)

```

-   Note that the number of rows in the joined result is greater than the target.
-   **This is because some cycle hire stations in cycle_hire have multiple matches in** `cycle_hire_osm`.
    -   our method generated multiple candidate points to be coupled with the official data.
-   Use NON-SPATIAL [aggregation methods]{.underline}:
    -   Take the capacity mean of the candidates and assign that to the corresponding point in the official data.

```{r}
aux = aux %>% 
  group_by(id) %>% 
  summarize(capacity = mean(capacity))
nrow(aux) == nrow(cycle_hire)
#> [1] TRUE
```

```{r}
plot(cycle_hire_osm["capacity"], main="actual capacity")
```

```{r}
plot(aux["capacity"], main= "estimated capacity")
```

------------------------------------------------------------------------

## NEXT CLASS: Spatial Aggregation and Raster operations!
